Reconstructing contact network parameters from viral phylogenies

### 2016-07-06 doc/paper/abstract.tex
Add paper and tex

Models of the spread of disease in a population often make the simplifying
assumption that the population is homogeneously mixed, or is divided into
homogeneously mixed compartments. However, human populations have complex
structures formed by social contacts, which can have a significant influence on
the rate of epidemic spread. Contact network models capture this structure by
explicitly representing each contact which could possibly lead to a
transmission. We developed a method based on approximate Bayesian
computation (ABC) for estimating structural parameters of the contact network
underlying an observed viral phylogeny. The method combines adaptive sequential
Monte Carlo for ABC, Gillespie simulation for propagating epidemics though
networks, and a kernel-based tree similarity score. We used the method to fit
the Barab\'{a}si-Albert network model to simulated transmission trees, and also
applied it to viral phylogenies estimated from \del{five}\add{ten} published
HIV sequence datasets. On simulated data, we found that the preferential
attachment power and the number of infected nodes in the network can often be
accurately estimated. On the other hand, the mean degree of the network, as
well as the total number of nodes, were not estimable with ABC. We observed
substantial heterogeneity in the parameter estimates on real datasets, with
point estimates for the preferential attachment power ranging from 0.06 to
1.05. These results underscore the importance of considering contact structures
when performing phylodynamic inference. Our method offers the potential to
quantitatively investigate the contact network structure underlying viral
epidemics.

>>>>

Models of the spread of disease in a population often make the simplifying
assumption that the population is homogeneously mixed, or is divided into
homogeneously mixed compartments. However, human populations have complex
structures formed by social contacts, which can have a significant influence on
the rate of epidemic spread. Contact network models capture this structure by
explicitly representing each contact which could possibly lead to a
transmission. We developed a method based on approximate Bayesian
computation (ABC) for estimating structural parameters of the contact network
underlying an observed viral phylogeny. The method combines adaptive sequential
Monte Carlo for ABC, Gillespie simulation for propagating epidemics though
networks, and a kernel-based tree similarity score. We used the method to fit
the Barab\'{a}si-Albert network model to simulated transmission trees, and also
applied it to viral phylogenies estimated from \del{five}\add{ten} published
HIV sequence datasets. On simulated data, we found that the preferential
attachment power and the number of infected nodes in the network can often be
accurately estimated. On the other hand, the mean degree of the network, as
well as the total number of nodes, were not estimable with ABC. \del{We
observed substantial heterogeneity in the parameter estimates on real datasets,
with point estimates for the preferential attachment power ranging from 0.06 to
1.05.} \add{We observed sub-linear preferential attachment (PA) power in all
datasets, as well as higher PA power in networks of injection drug users.}
These results underscore the importance of considering contact structures when
performing phylodynamic inference. Our method offers the potential to
quantitatively investigate the contact network structure underlying viral
epidemics.

### 2016-07-06 doc/paper/changes.tex
Add paper and tex

\signature{Rosemary M. McCloskey}
\address{Centre for Excellence in HIV/AIDS \\ 608-1081 Burrard Street \\ Vancouver, BC, Canada V6Z 1Y6}
\begin{document}
\begin{letter}{ }

@@

  \item It was pointed out to us by a colleague that the posterior mean is a
    more conventionally used Bayesian estimator than the maximum \textit{a
    posteriori} (MAP) value. Therefore, we have replaced the MAP point
    estimates by posterior means. Qualitatively, the results are not much
    different, except that it is more obvious that $m$ and $N$ are not
    identifiable with this method. 
  \item We have replaced some ad hoc statistical tests in the analysis of the
    simulation results with a $p$-value corrected generalized linear model, and
    rewritten the relevant section of the results accordingly.
  \item We have found an analysis of the Barab\'asi-Albert model in the physics
    literature (Krapivsky et al. 2000) which demonstrates that the degree
    distribution is only a power law when $\alpha = 1$. We acknowledge this but
    do not believe it detracts from the validity of our results, especially in
    light of a paper by de Blasio et al. (2007) which found sub-linear
    preferential attachment through a population-level survey. We have added to
    the supplemental materials a series of power law fits to the degree
    distributions of BA networks with $\alpha \neq 1$ (Figure TODO), which
    shows that, although the power law is cleary an imperfect fit, the slope
    is still reasonably representative of the true distribution.
  \item We did some additional literature searching and found five more
    published HIV datasets that seemed to be appropriate for our method. For
    clarity, the real HIV datasets are now referred to by risk group and
    geographic location, rather than by author (eg. MSM/Beijing instead of Wang
    et al.). The highest $\alpha$ values were associated with IDU networks,
    coroborating the assertion of Dombrowski et al. (2013) that IDU networks
    exhibit the scale-free property more strongly than sexual networks do. 
  \item Three of these new datasets had sequenced two HIV genes, which we
    analysed separately. We found that estimates of $\alpha$ were very robust
    to the choice of gene; $I$ was less robust, with \textit{env} genes tending
    to lead to higher estimated values.
  \item Figure 5 (formerly Figure 4) now shows 50\% HPD intervals (similar to
    interquartile ranges) in addition to 95\% HPD intervals.
  \item We discovered that our use of the term ``kernel-ABC'' to refer to ABC 
    with a kernel for the distance function is in conflict with another usage,
    namely an expression of Bayes' rule in terms of kernel means (Fukumizu et
    al. ``Kernel Bayes' rule.'' 2011). We therefore simply use the term ``ABC''
    to refer to our method.
  \item The discussion has been divided into subsections to make it easier to
    follow.
\end{itemize}

@@

We are grateful that the reviewer finds our work to be of some interest, and we
acknowledge that our ability to precisely resolve numerical values of the BA
parameters is limited. We believe that our simulation results show that some
network parameters (specifically $\alpha$ in our case) can be at least roughly
estimated from the information contained in tree shapes.

We have included several new analyses that attempt to disambiguate true
features of the posterior distribution from artefacts of a poor ABC
approximation. First, we used ABC to estimate marginal posterior distributions
when some combinations of parameters were held fixed to known values.
Estimation error which is removed during marginal estimation is likely to be
due to confounding between model parameters; error which remains may be due to
either poor resolution or truely diffuse posteriors. The accuracy of the
estimates was improved about 50\% in marginal estimation.

Second, we examined the possible Monte Carlo error by increasing the resolution
of the ABC approximation (more particles, more simulated datasets per particle,
and more iterations). The results were nearly identical, TODO

Finally, three of the newly included HIV datasets included both \textit{gag}
and \textit{env} sequences which we analysed separately. The estimates of
$\alpha$ for the two genes were nearly identical for each of the datasets,
which seems unlikely if the ABC approximations were very poor.

@@

We apologise for this misunderstanding due to our poor choice of words. An SVR
is in fact a regressor, not a classifier, so the reported $R^2$ were of true
values against predictions. However, given that the point of this experiment
was to demonstrate identifiability by categorizing tree shapes into
predetermined values, we redid the analysis using SVMs (which are true
classifiers) instead of SVRs and linear regression. We now report the accuracy
of the categorical classification, and figures 1 and S3-S6 have been revised
accordingly. Some additional variation among meta-parameters and epidemic
scenarios is observed, but the overall results are similar and the best choice
of meta-parameters is the same.

@@

The method we used to estimate transmission trees simply reroots and adjusts
the branch lengths of an estimated viral phylogeny, which is always binary as
long as there are no duplicated sequences. We have added the following
clarifications in the methods section: (1) ``Duplicated sequences were removed
with BioPython'' (2) ``Due to the removal of duplicated sequences, all
estimated transmission trees were fully binary.''

@@

  TODO

@@

We agree with this assessment of our visual presentation of the results. We
have reduced the number of posterior plots shown to one in the text and two in
the supplemental materials. Figure 2, which formerly represented each
simulation by an individual point, has been replaced by a box plot showing the
$\alpha$ and $I$ estimates stratified by each of the other parameters. A
similar boxplot for $m$ and $N$, which were not identifiable with ABC, has been
added to the supplemental materials.

@@

  The Discussion focusses around model mis-specification, which is very
  relevant (and to which the inference isn't that robust, suggesting that it
  might also not be robust to other things like dynamics, clustering,
  age-related mixing etc). But another question is how much of the signal you
  would expect to find in this simulation strategy, and how you would expect to
  see it using these kinds of tree comparisons. I don't have much intuition
  about the kernel but the Sackin imbalance will presumably be lower where
  there are very high-degree nodes, and higher where there are long
  transmission chains. The nLTT will depend on the epidemic curves, presumably.
\end{quote}

The identifiability of the BA parameters is of course a critical prerequisite
to the kind of inference we are trying to perform. Indeed, the objective of the 
analysis with classifiers was to investigate identifiability, and we apologise
for not making this clearer. The reviewer's intuition is correct -- we do
expect imbalance to be sensitive to the degree distribution (related to
$\alpha$) and the nLTT to be sensitive to the epidemic curves (related to $I$
and $N$). This is reflected in the performance of the respective classifiers:
Sackin's index does better than the nLTT at classifying $\alpha$, but the nLTT
is better at classifying $I$ and $N$.

@@

  L45-46: “…and in epidemics of differing size (500, 1000 and 2000)”. Unless
  I’m missing something this is just a restatement of what’s already in this
  paragraph about I, hence repetitive. 
\end{quote}

We apologise for the misunderstanding. Because $I$ is not a structural
parameter of the network, we attempted to classify each of $\alpha$, $N$, and
$m$ under several different fixed values of $I$. We have rephrased this as
follows: ``For the structural parameters $\alpha$, $m$, and $N$, the
experiments were repeated with three different fixed values of $I$ (500, 1000,
and 2000).''

@@

  P7 \\
  L20-24: It’s slightly surprising to me that the time trees were constructed
  using “fast” methods when the datasets are not especially large. Were there
  problems in obtaining trees with other algorithms? While I do not insist that
  every analysis be redone using alternative methods, it would be reassuring to
  know that the results were consistent, in at least one example, if
  reconstruction were done with, for example, RAxML+LSD, or by taking the BEAST
  MCC tree. Similarly, are results similar if analysis is performed on several
  bootstraps or posterior samples?
\end{quote}

Although they are very approximate methods, our experience has been that using
approximate maximum likelihood phylogeny reconstruction with root-to-tip
regression does not result in large errors in phylodynamic inference (Poon
2015). We reanalysed the datasets with the highest and lowest estimated
$\alpha$ (Zetterberg et al. 2004 and Cuevas et al. 2009) using a BEAST MCC tree
from a run with simple models of evolution and population size. The results
were TODO

With respect to the suggestion to analyse bootstrap replicates, we believe that
the three new datasets containing both \textit{gag} and \textit{env} data fill
the need for within-dataset replication.

@@

  L32: “Estimated correctly” – i.e. the posterior mode was correct?
\end{quote}

@@

  P13 \\
  Figure 3, legend: I’m guessing that the shaded polygon is the HPD region,
  rather than the “true value”?
\end{quote}

@@

  P15 \\
  L38-40: While I understand the point, it’s surely quite unlikely that the
  network will be identical. Better to say that the number of infected nodes
  with new children is small, and hence there is little difference in the
  outcomes of the simulations.
\end{quote}

We struggled with the phrasing of this particular point and are happy to have
recieved some guidance. TODO

@@

  L41-42: “We note also that our accurate estimates of I may have been
  influenced by this prior”. Couldn’t you check?
\end{quote}

This issue was also addressed by performing marginal estimation. If the
accurate estimates of $I$ were solely due to the prior, they would have gotten
worse, not better, when $N$ was fixed and the prior on $I$ was truly uniform.
We make this point in the discussion: ``when $I$ was estimated marginally with
fixed $N$, the accuarcy of the estimate improved even though there was no
longer any extra prior mass on low $I$ values.''

@@

We had been using the term ``prevalence'' interchangably with $I$, but since
this is ambiguous we have replaced all instances of ``prevalence'' with $I$
except for those which do not refer to $I$ explicitly.

@@

  A point I’d like to see at least discussed – presumably if some iterations
  add nodes with one connection and others add them with two, you’d get cycles,
  but not as many as you’d expect with m=2. Is this worth considering in the
  question of why the binary choice to include m=1 is so problematic? Would it
  be better if m was drawn from a discrete distribution at every step?
\end{quote}

This is a valid point, and in fact drawing $m$ from a distribution is an
approach we considered when we were performing this work. However, since this
is our first attempt to fit network models, we thought it best to stick with
a simple, well-known model rather than defining our own parameters. We did
perform a small experiment allowing heterogeneity among nodes, but in $\alpha$
rather than in $m$, and recovered the average of the two $\alpha$ values.

(TODO: We
have added to the discussion the point that this binary choice is sub-optimal
and suggested this enhancement for future work.)

\begin{quote}
  Supplement:
  Figure S1 legend: What is the shaded area?
\end{quote}

>>>>

\signature{Rosemary M. McCloskey}
\address{BC Centre for Excellence in HIV/AIDS \\ 608-1081 Burrard Street \\ Vancouver, BC, Canada V6Z 1Y6}
\begin{document}
\begin{letter}{ }

@@

  \item It was pointed out to us by a colleague that the posterior mean is a
    more conventionally used Bayesian estimator than the maximum \textit{a
    posteriori} (MAP) value. Therefore, we have replaced the MAP point
    estimates by posterior means. Qualitatively, the results are not much
    different.
  \item We have replaced some ad hoc statistical tests in the analysis of the
    simulation results with a multiple test corrected generalized linear model,
    and revised the relevant section of the results accordingly.
  \item We have found an analysis of the Barab\'asi-Albert model in the physics
    literature (Krapivsky et al. 2000) which demonstrates that the degree
    distribution is only a power law when $\alpha = 1$. We acknowledge this in
    the revised manuscript but do not believe it detracts from the validity of
    our results, especially in light of a paper by de Blasio et al. (2007)
    which found sub-linear preferential attachment through a population-level
    survey. We have added to the supplemental materials a series of power law
    fits to the degree distributions of BA networks with $\alpha \neq 1$
    (Figure S11), which shows that, although the power law is clearly an
    imperfect fit, the slope is still reasonably representative of the true
    distribution.
  \item We did some additional literature searching and found five more
    published HIV datasets that seemed to be appropriate for our method. For
    clarity, the real HIV datasets are now referred to by risk group and
    geographic location, rather than by author (eg. MSM/Beijing instead of Wang
    et al.). The highest $\alpha$ values were associated with IDU networks,
    corroborating the assertion of Dombrowski et al. (2013) that IDU networks
    exhibit the scale-free property more strongly than sexual networks do. 
  \item Two of these new datasets had sequenced two different HIV genes
    (\textit{gag} and \textit{env}), which we analysed separately. We found
    that estimates of $\alpha$ were very robust to the choice of gene. $I$ was
    less robust -- in one of the two datasets, the \textit{env} sequences led
    to a noticably higher estimated value than the \textit{gag} sequences.
  \item We discovered that our use of the term ``kernel-ABC'' to refer to ABC 
    with a kernel for the distance function is in conflict with another usage,
    namely an expression of Bayes' rule in terms of kernel means (Fukumizu et
    al. ``Kernel Bayes' rule.'' 2011). We therefore simply use the term ``ABC''
    to refer to our method.
  \item Figure 5 (formerly Figure 4) now shows 50\% HPD intervals (similar to
    interquartile ranges) in addition to 95\% HPD intervals.
  \item The discussion has been divided into subsections to make it easier to
    follow.
\end{itemize}

@@

We are grateful that the reviewer finds our work to be of some interest. Even
in the relatively simple scenario we studied (susceptible-infected epidemics
over Barab\'asi-Albert networks), the process leading from a network model to a
transmission tree is complex. To calculate the likelihood of an observed
transmission tree exactly, it would be necessary to sum over all possible
contact networks the model could generate \emph{and} all possible epidemics
over each contact network. It is therefore unsurprising that the posterior
distributions are diffuse. In light of this, the level of parameter
identifiability we observed is better than one might expect \textit{a priori}.
We are confident that our parameter estimates, especially of $\alpha$, are
truly representative of the posterior distributions and not simply artifacts of
a poor ABC approximation. Several new analyses have been included to
demonstrate this.

First, we used ABC to estimate marginal posterior distributions when some
combinations of parameters were held fixed to known values. Estimation error
which is removed during marginal estimation is likely to be due to confounding
between model parameters. The accuracy of the estimates was improved about 50\%
in marginal estimation (Figure S8), implying that at least half of the observed
inaccuracy stems from parameter correlations in the posterior. Although we have
no way to prove it, it seems highly likely that even the marginal posterior
distributions are diffuse given the complexity of the simulation process noted
above. If that is the case, even more of the error may be attributed to true
posterior features.

Second, we examined the possible Monte Carlo error by increasing the resolution
of the ABC approximation (twice as many particles, twice as many simulated
datasets per particle, and roughly 1.5 times as many iterations). The results
were comparable (Figure S9), which demonstrates that Monte Carlo error from the
SMC algorithm is likely not responsible for observed estimation errors.

Finally, two of the newly included HIV datasets included both \textit{gag}
and \textit{env} sequences which we analysed separately. The estimates of
$\alpha$ for the two genes were very similar for each of the datasets (Figure
S12), which seems unlikely if the ABC approximations were very poor. We also
reanalyzed two of the datasets with a different phylogenetic reconstruction
method, and again found very similar estimates of $\alpha$ (Figure S13). The
estimates of $I$ were less robust to both of these variables, but remained in
same sub-2000 region of the range in both cases. In the revised discussio, we
acknowledge that $I$ was overestimated in simulations and offer the possible
explanation that $I$ is only identifiable up to the epidemic phase it
represents (\textit{i.e.} exponential growth vs. linear expansion vs. slow
growth, Figure S15).

@@

We apologize for this misunderstanding due to our poor choice of words. An SVR
is in fact a regressor, not a classifier, so the reported $R^2$ were of true
values against predictions. However, given that the point of this experiment
was to demonstrate identifiability by categorizing tree shapes into
predetermined values, we redid the analysis using SVMs (which are true
classifiers) instead of SVRs and linear regression. We now report the accuracy
of the categorical classification, and figures 1 and S3-S6 have been revised
accordingly. Some additional variation among meta-parameters and epidemic
scenarios is observed, but the overall results are very similar.

@@

The method we used to estimate transmission trees simply reroots and adjusts
the branch lengths of an estimated viral phylogeny, which is always binary as
long as there are no duplicated sequences. We have added the following
clarifications in the methods section: (1) ``Duplicated sequences were removed
with BioPython''; and (2) ``Due to the removal of duplicated sequences, all
estimated transmission trees were fully binary.''

@@

We have substantially expanded the discussion of the synthetic data results,
including this point about $m$. The weaker identifiability in lower $\alpha$
values may be related to the degree distributions of the networks, and the
upward biases in $I$ and $N$ are discussed in relation to the epidemic
trajectory of the SI model. Our intuition about $m$ is weaker, but we include
this point by the reviewer and hypothesize that, as reviewer 2 suggests, a
model which allows heterogeneity in $m$ may offer better identifiability.

@@

We agree with this assessment of our visual presentation of the results. We
now only show a single set of marginal posterior distributions in the text.
Figure 2, which formerly represented each simulation by an individual point,
has been replaced by a box plot showing the $\alpha$ and $I$ estimates
stratified by each of the other parameters. A similar boxplot for $m$ and $N$,
which were non- and weakly-identifiable respectively with ABC, has been added
to the supplemental materials (Figure S7).

@@

  The Discussion focuses around model mis-specification, which is very
  relevant (and to which the inference isn't that robust, suggesting that it
  might also not be robust to other things like dynamics, clustering,
  age-related mixing etc). But another question is how much of the signal you
  would expect to find in this simulation strategy, and how you would expect to
  see it using these kinds of tree comparisons. I don't have much intuition
  about the kernel but the Sackin imbalance will presumably be lower where
  there are very high-degree nodes, and higher where there are long
  transmission chains. The nLTT will depend on the epidemic curves, presumably.
\end{quote}

The identifiability of the BA parameters is of course a critical prerequisite
to the kind of inference we are trying to perform. Indeed, the objective of the 
analysis with classifiers was to investigate identifiability, and we apologize
for not making this clearer. The reviewer's intuition is correct -- we do
expect imbalance to be sensitive to the degree distribution (related to
$\alpha$) and the nLTT to be sensitive to the epidemic curves (related to $I$
and $N$). This is reflected in the performance of the respective classifiers:
Sackin's index does better than the nLTT at classifying $\alpha$, but the nLTT
is better at classifying $I$ and $N$.

@@

  L45-46: ``\ldots and in epidemics of differing size (500, 1000 and 2000)''.
  Unless I'm missing something this is just a restatement of what's already in
  this paragraph about I, hence repetitive. 
\end{quote}

We apologize for the misunderstanding. Because $I$ is not a structural
parameter of the network, we attempted to classify each of $\alpha$, $N$, and
$m$ under several different fixed values of $I$. We have rephrased this as
follows: ``For the structural parameters $\alpha$, $m$, and $N$, the
experiments were repeated with three different fixed values of $I$ (500, 1000,
and 2000).''

@@

  P7 \\
  L20-24: It's slightly surprising to me that the time trees were constructed
  using ``fast'' methods when the datasets are not especially large. Were there
  problems in obtaining trees with other algorithms? While I do not insist that
  every analysis be redone using alternative methods, it would be reassuring to
  know that the results were consistent, in at least one example, if
  reconstruction were done with, for example, RAxML+LSD, or by taking the BEAST
  MCC tree. Similarly, are results similar if analysis is performed on several
  bootstraps or posterior samples?
\end{quote}

Although they are very approximate methods, our experience has been that using
approximate maximum likelihood phylogeny reconstruction with root-to-tip
regression does not result in large errors in phylodynamic inference (Poon
2015). We reanalysed two datasets (Niculescu et al. 2015 and Cuevas et al.
2009) using RAxML and LSD. The results for $\alpha$ were nearly identical, but
$I$ was not robust to the reconstruction method -- RAxML+LSD produced higher
$I$ estimates for both datasets (Figure S13). This result has been added to the
discussion of both parameters.

Rather than analyze bootstrap replicates, we used the two new datasets
containing both \textit{gag} and \textit{env} sequences for within-dataset
replication.

@@

  L32: ``Estimated correctly'' -- i.e. the posterior mode was correct?
\end{quote}

@@

  P13 \\
  Figure 3, legend: I'm guessing that the shaded polygon is the HPD region,
  rather than the ``true value''?
\end{quote}

@@

  P15 \\
  L38-40: While I understand the point, it's surely quite unlikely that the
  network will be identical. Better to say that the number of infected nodes
  with new children is small, and hence there is little difference in the
  outcomes of the simulations.
\end{quote}

We struggled with the phrasing of this particular point and are happy to have
received some guidance. The statement now reads as follows: ``If $I$ is small
relative to $N$, very few of the infected nodes will gain any new neighbours.
Thus, the outcome of a second simulation on the same network will likely be
very similar.''

@@

  L41-42: ``We note also that our accurate estimates of I may have been
  influenced by this prior''. Couldn't you check?
\end{quote}

This issue was addressed by performing marginal estimation. If the accurate
estimates of $I$ were solely due to the prior, they would have gotten worse,
not better, when $N$ was fixed and the prior on $I$ was truly uniform. We make
this point in the discussion: ``when $I$ was estimated marginally with fixed
$N$, the accuracy of the estimate improved even though there was no longer any
extra prior mass on low $I$ values.''

@@

We had been using the term ``prevalence'' interchangeably with $I$, but since
this is ambiguous we have replaced all instances of ``prevalence'' with $I$
except for those which do not refer to $I$ explicitly.

@@

  A point I'd like to see at least discussed -- presumably if some iterations
  add nodes with one connection and others add them with two, you'd get cycles,
  but not as many as you'd expect with m=2. Is this worth considering in the
  question of why the binary choice to include m=1 is so problematic? Would it
  be better if m was drawn from a discrete distribution at every step?
\end{quote}

This is a valid point, and in fact drawing $m$ from a distribution is an
approach we considered when we were performing this work. However, since this
is our first attempt to fit network models, we thought it best to stick with
a simple, well-known model rather than defining our own parameters. We did
perform a small experiment allowing heterogeneity among nodes, but in $\alpha$
rather than in $m$, and recovered the average of the two $\alpha$ values. We
have outlined the difficulties of the binary choice of $m$ in the discussion of
the real data results: ``\ldots the BA model does not allow any nodes with
degree 1 when $m \geq 2$. For this reason, the choice of whether to allow $m =
1$ in the prior is problematic, as we must choose between an unrealistic
topology (no cycles) and an unrealistic minimum degree.''

  Supplement: \\
  Figure S1 legend: What is the shaded area?
\end{quote}
